Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument

### 2014-10-15 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
file added

\title[Modal Collapse]
 {The Ontological Modal Collapse \\ 
 as a Collapse of the Square of Opposition}

%----------Author 1
\author[Benzm\"uller]{Christoph Benzm\"uller}

@@

%----------classification, keywords, date
\subjclass{
Prim. 03A02;  % Philosophical aspects of logic and foundations
Sec. 68T02 % Artificial Intelligence
}

\keywords{Modal Logics, Higher-Order Logics, Ontological Argument}

\date{August 30, 2014}
%----------additions
\dedicatory{ }
%%% ----------------------------------------------------------------------

\begin{abstract}
The \emph{modal collapse} that afflicts G\"odel's modal ontological 
argument for God's existence is discussed from the perspective of the 
modal square of opposition. Furthermore, a computer-assisted verification 
of the claims that the emendations by Anderson and by Frode are immune to 
the modal collapse is presented.
\end{abstract}

%%% ----------------------------------------------------------------------
\maketitle
%%% ----------------------------------------------------------------------
%\tableofcontents
\section{Introduction}

Attempts to prove the
existence (or non-existence) of God by means of abstract, ontological
arguments are an old tradition in western philosophy, with contributions by several prominent philosophers, including St. Anselm of
Canterbury, Descartes and Leibniz. Kurt G{\"o}del studied and further improved this argument, bringing it to a mathematically more precise form, as a chain of axioms, lemmas and theorems in a modal logic \cite{GoedelNotes,ScottNotes}, as shown in Fig. \ref{fig:scott}.

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[L1] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[L2] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Scott's version of G\"odel's ontological argument \cite{ScottNotes}.\label{fig:scott}} 
\end{figure}

% Ontological arguments, for or against the existence of God,
% illustrate well an essential aspect of metaphysics: some (necessary) facts
% for our existing world are deduced by purely a priori, analytical means from some
% abstract definitions and axioms. % Contingent truths are to be
% distinguished from necessary truths.

G\"{o}del defines God (see Fig.~\ref{fig1}) as a being who possesses all \emph{positive}
properties.  He does not extensively discuss what positive properties
are, but instead he states a few reasonable (but debatable) axioms
that they should satisfy.  
The overall idea of G{\"o}del's proof is in the tradition of Anselm's
argument, who defined God as some entity of which nothing greater can be
conceived. Anselm argued that existence in the actual world would
make such an assumed being even greater; hence, by definition God must
exist. G{\"o}del's ontological argument is clearly related to this
reasoning pattern. However, it also tries to fix some fundamental
weaknesses in Anselm's work. For example, G{\"o}del explicitly proves
that God's existence is possible, which has been a basic assumption for
Anselm. Because of this, Anselm's argument has been
criticized as incomplete by Leibniz. Leibniz claimed that the
assumption should be derivable from the definition of God as a perfect
being and from the notion of perfection.  G{\"o}del's proof 
addresses this critique, and it also addresses Kant's objection that existence should not be treated as a predicate.

Nevertheless, G{\"o}del's work
still leaves room for criticism. In particular, his axioms are so
strong that they entail a \emph{modal collapse} 
\cite{ToDo:sobel1987?,sobel2004logic}: 
everything that is the case is so necessarily. 
There has been an impressive body of recent and ongoing
work (cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,ToDo:FrodeAndHajek,ContemporaryBibliography} and the references therein)
proposing solutions for the modal collapse. 
The goal of this paper is to present a computer-assisted analysis of 
the solutions proposed by Anderson and by Frode. The technique enabling this analysis is the embedding of quantified modal logics into higher-order logics \cite{J23,B9,C36}, for which automated theorem provers exist \cite{ToDo:leo,sattalax,isabelle,coq}. This technique has already been succesfully employed in the verification and reconstruction of G\"odel's proof \cite{todo:all our previous papers}, and a detailed mathematical description of the technique is available in \cite{todo:ECAI}.

\section{The Modal Collapse}

A crucial step of the ontological argument is the claim that if God's existence is possible, then it is necessary. This is Lemma L2 in G\"odel's proof. In the modal square of opposition (Fig. \ref{ToDo}), this is an unusual situation in which the \textbf{I} corner implies and entails the \textbf{A} corner, for the particular case when $\phi$ is $\exq x G(x)$. G\"odel's proof shows that his axioms are strong enough to invert the direction of entailment for the particular $\phi$ at issue. The question is whether the axioms are then not too strong, also allowing the inverted entailment for arbitrary $\phi$. That is essentially the question asked by Sobel \cite{Sobel1987}; and his proof of the modal collapse (MC) answers it affirmatively. The modal collapse is a collapse of the modal square: it shows that, assuming G\"odel's axioms, the subcontraries entail the contraries.

>>>>

\title[Modal Collapse]
 {Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument}

%----------AUTHOR 1
\author[Benzm\"uller]{Christoph Benzm\"uller}

@@

%----------classification, keywords, date
% \subjclass{
% % Prim. 03A02;  % Philosophical aspects of logic and foundations
% % Sec. 68T02 % Artificial Intelligence 
% }

%\keywords{Ontological Argument, Modal Collapse}

\date{\today}
%----------additions
\dedicatory{ }
%%% ----------------------------------------------------------------------

% \begin{abstract}
% The \emph{modal collapse} that afflicts G\"odel's modal ontological 
% argument for God's existence is discussed from the perspective of the 
% modal square of opposition. Furthermore, a computer-assisted verification 
% of the claims that the emendations by Anderson and by Frode are immune to 
% the modal collapse is presented.
% \end{abstract}

%%% ----------------------------------------------------------------------
\maketitle
%%% ----------------------------------------------------------------------
%\tableofcontents
%\section{Introduction}

\section{The Modal Collapse}
The axioms in G\"odel's ontological argument \cite{GoedelNotes,ScottNotes} imply what is
called the modal collapse \cite{Sobel1987,SobelBook} (formula $\varphi \rightarrow \Box \varphi$
holds for any formula $\varphi$ and not just for $\exists x. God(x)$
as intended). This fact, which has recently been confirmed with
higher-order automated theorem provers \cite{}, has led to strong
criticism in the research community, and it has stimulated further
adaptations of G\''odel's argument with the aim to remedy the situation
\cite{}.

In recent work we have formalized and automated the proposed adaptions
of Anderson \cite{}, Bjordahl \cite{} and Fuhrmann \cite{}.

\pagebreak

 crucial step of the ontological argument is the claim that if God's existence is possible, then it is necessary. This is Lemma L2 in G\"odel's proof. In the modal square of opposition (Fig. \ref{ToDo}), this is an unusual situation in which the \textbf{I} corner implies and entails the \textbf{A} corner, for the particular case when $\phi$ is $\exq x G(x)$. G\"odel's proof shows that his axioms are strong enough to invert the direction of entailment for the particular $\phi$ at issue. The question is whether the axioms are then not too strong, also allowing the inverted entailment for arbitrary $\phi$. That is essentially the question asked by Sobel \cite{Sobel1987}; and his proof of the modal collapse (MC) answers it affirmatively. The modal collapse is a collapse of the modal square: it shows that, assuming G\"odel's axioms, the subcontraries entail the contraries.
